---
title: "DannyNgo_Assignment_4"
author: "Danny Ngo"
date: "2022-11-11"
output: html_document
---

### Import Libraries
```{r}
library(vegan)
library(ggplot2)
```

### Generate Community Matrix for Species vs Communities
```{r}
set.seed(2)
community1 = matrix(sample(0:250,300,replace=T),nrow=10,dimnames=list(paste("community",1:10,sep=""),paste("sp",1:30,sep="")))
community2 = matrix(sample(50:200,300,replace=T),nrow=10,dimnames=list(paste("community",11:20,sep=""),paste("sp",1:30,sep="")))
community_matrix <- rbind(community1,community2)
```

### Generate Metadata depicting the treatments
```{r}
m1 <- as.data.frame(list(paste("community",1:10,sep=""), paste("T1")))
colnames(m1) <- c("Communities", "Treatment")
m2 <- as.data.frame(list(paste("community",11:20,sep=""), paste("T2")))
colnames(m2) <- c("Communities", "Treatment")
metadata <- rbind(m1,m2)
```

### 1. Calculate the Shannon diversity index for all samples using the vegan package.
```{r}
raremax <- min(rowSums(community_matrix))
rarefied_data <- rrarefy(community_matrix, raremax)
shannon <- vegan::diversity(rarefied_data, index = "shannon")
diversity_tbl <- cbind(metadata, shannon)
diversity_tbl
```

### 2. Using the Shannon diversity measurements obtained in Step 1, make a graph with a boxplot for each treatment - customize the plot as you wish.
```{r}
plot <- ggplot(diversity_tbl, aes(x=Treatment, y=shannon)) +
  geom_boxplot(outlier.shape=4) +
  theme(axis.text.x = element_text(hjust = 1)) +
  labs(y="Shannon diversity index") +
  geom_jitter(aes(color=shannon)) +
  theme(legend.position = "none")
plot
```

### 3. Using the Shannon diversity measurements obtained in Step 1, compare the treatments using a t-test. There is no need to verify data normality.
```{r}
t_test <- t.test(shannon ~ Treatment, diversity_tbl)
t_test
```

### 4. Make an NMDS using the metaMDS function from the vegan package. Use the “bray” distance. Plot the NMDS using ggplot2 with a different color for each treatment. There is no need to add the arrows in the ordination. Plot a graph with the communities only. See the video on Week 9 as an example. 
```{r}
ordination <- metaMDS(community_matrix, "bray")
ef <- envfit(ordination, community_matrix, nperm = 999)
ef.adj <- ef 
pvals.adj <- p.adjust(ef$vectors$pvals, method = "fdr") 
ef.adj$vectors$pvals <- pvals.adj 
ef.adj

site <- as.data.frame(scores(ordination, display = "sites")) 
site <- cbind(site,Treatment=metadata$Treatment)

nmds <- ggplot(site, aes(x = NMDS1, y = NMDS2)) + 
  geom_point(aes(NMDS1, NMDS2, color = factor(site$Treatment), 
                 shape = factor(site$Treatment)), size = 6) + 
  labs(colour = "Treatment", shape = "Treatment") 
nmds
```

### 5. Run a PERMANOVA analysis to verify the difference between treatments.
```{r}
clr <- decostand(community_matrix, method = "rclr")
dist <- vegdist(clr, method = "euclidean")
permanova = adonis2(dist ~ Treatment, data = site, permuations = 999)
permanova
```