---
title: "DannyNgo_Grad_Assignment"
author: "Danny Ngo"
date: "2022-12-05"
output: html_document
---

# Introduction
Dataset of total RNA in lung tissue was obtained through quantitative PCR from a paper testing SARS-CoV-2 breakthrough infections in K18-hACE2 transgenic mice. Waning vaccine protection was evaluated by challenging non-boosted mice (Mock) and mice boosted with human COVID-19 mRNA post-vaccination sera (Post-vac) with Delta, Kappa, and NY (614G) variants of SARS-CoV-2. Differential gene expression was assessed through volcano plots allowing quick visualization of genes with significant fold change.

Citation:

Kwon, H. J., Kosikova, M., Tang, W., Ortega-Rodriguez, U., Radvak, P., Xiang, R., Mercer, K. E., Muskhelishvili, L., Davis, K., Ward, J. M., Kosik, I., Holly, J., Kang, I., Yewdell, J. W., Plant, E. P., Chen, W. H., Shriver, M. C., Barnes, R. S., Pasetti, M. F., Zhou, B., … Xie, H. (2022). Enhanced virulence and waning vaccine-elicited antibodies account for breakthrough infections caused by SARS-CoV-2 delta and beyond. iScience, 25(12), 105507. https://doi.org/10.1016/j.isci.2022.105507

Dataset: GSE193839 https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE193839

***

### Import Libraries
```{r}
library(tidyverse)
library(ggplot2)
library(ggrepel)
```

### Import Dataset
```{r}
data <- read.delim("E:/School UF/2022 Fall - MCB6937 - R for Life Sciences/Grad Assignment/GSE193839_non-normalized_data.txt", header=FALSE)
head(data)
```

### Format Dataset
```{r}
df <- na_if(data[3:92,1:40], "No Ct") # Replace "No Ct" values with "NA"
colnames(df) <- df[1,1:40] # set column headers to first row
df = df[-1,-1] # remove row and column header in data frame
df <- data.frame(apply(df, 2, function(x) as.numeric(as.character(x)))) # convert data class to Numeric
rownames(df) <- data[4:92,1] # set row headers
```

### Set up t-test Function to Calculate Difference Between Control Group and Test Group
```{r}
t_test_function <- function(df, control, test_group) {
  x = df[control]
  y = df[test_group]
  results = t.test(x, y)
  results$p.value # output vector p.value
}
```

### Apply t-test and Calculate Fold Change for Test Group 1 (NY_Mock)
```{r}
p_value = apply(df, 1, t_test_function, control = c(1:3), test_group = c(4:9))
log2_df = log2(df)
control = apply(log2_df[,1:3], 1, mean, na.rm=TRUE)
test = apply(log2_df[,4:9], 1, mean, na.rm=TRUE)
LogFC <- control - test
new_data <- data.frame(ID = data[4:92,1], LogFC, p_value)
new_data <- new_data %>%
  mutate(Expression = dplyr::case_when(LogFC >= 0 & p_value <= 0.05 ~ "Up-regulated",
                                       LogFC <= 0 & p_value <= 0.05 ~ "Down-regulated",
                                       TRUE ~ "Unchanged"))
top_id <- new_data %>% # add gene names
  filter(Expression == "Up-regulated" | Expression == "Down-regulated")
```

### Create Volcano Plot for Test Group 1 (NY_Mock)
```{r}
theme_set(theme_bw())

p1 <- ggplot(new_data, aes(x = LogFC, y = -log(p_value, 10))) +
  geom_point(aes(color = Expression)) +
  ggtitle("Gene Expression in NY_Mock") +
  theme(plot.title = element_text(size = 24, hjust = 0.5)) +
  xlab(expression("log"[2]*"FC")) +
  ylab(expression("-log"[10]*"FDR")) +
  scale_color_manual(values = c("green", "grey50", "firebrick3")) +
  geom_label_repel(data = top_id, label = top_id$ID, size = 2, label.size = NA, fill = NA, max.overlaps = Inf)
p1
```

### Apply t-test and Calculate Fold Change for Test Group 2 (Kappa_Mock)
```{r}
p_value = apply(df, 1, t_test_function, control = c(1:3), test_group = c(10:15))
test = apply(log2_df[,10:15], 1, mean, na.rm=TRUE)
LogFC <- control - test
new_data <- data.frame(ID = data[4:92,1], LogFC, p_value)
new_data <- new_data %>%
  mutate(Expression = dplyr::case_when(LogFC >= 0 & p_value <= 0.05 ~ "Up-regulated",
                                       LogFC <= 0 & p_value <= 0.05 ~ "Down-regulated",
                                       TRUE ~ "Unchanged"))
top_id <- new_data %>%
  filter(Expression == "Up-regulated" | Expression == "Down-regulated")
```

### Create Volcano Plot for Test Group 2 (Kappa_Mock)
```{r}
p2 <- ggplot(new_data, aes(x = LogFC, y = -log(p_value, 10))) +
  geom_point(aes(color = Expression)) +
  ggtitle("Gene Expression in Kappa_Mock") +
  theme(plot.title = element_text(size = 24, hjust = 0.5)) +
  xlab(expression("log"[2]*"FC")) +
  ylab(expression("-log"[10]*"FDR")) +
  scale_color_manual(values = c("green", "grey50", "firebrick3")) +
  geom_label_repel(data = top_id, label = top_id$ID, size = 2, label.size = NA, fill = NA, max.overlaps = Inf)
p2
```

### Apply t-test and Calculate Fold Change for Test Group 3 (Delta_Mock)
```{r}
p_value = apply(df, 1, t_test_function, control = c(1:3), test_group = c(16:21))
test = apply(log2_df[,16:21], 1, mean, na.rm=TRUE)
LogFC <- control - test
new_data <- data.frame(ID = data[4:92,1], LogFC, p_value)
new_data <- new_data %>%
  mutate(Expression = dplyr::case_when(LogFC >= 0 & p_value <= 0.05 ~ "Up-regulated",
                                       LogFC <= 0 & p_value <= 0.05 ~ "Down-regulated",
                                       TRUE ~ "Unchanged"))
top_id <- new_data %>%
  filter(Expression == "Up-regulated" | Expression == "Down-regulated")
```

### Create Volcano Plot for Test Group 3 (Delta_Mock)
```{r}
p3 <- ggplot(new_data, aes(x = LogFC, y = -log(p_value, 10))) +
  geom_point(aes(color = Expression)) +
  ggtitle("Gene Expression in Delta_Mock") +
  theme(plot.title = element_text(size = 24, hjust = 0.5)) +
  xlab(expression("log"[2]*"FC")) +
  ylab(expression("-log"[10]*"FDR")) +
  scale_color_manual(values = c("green", "grey50", "firebrick3")) +
  geom_label_repel(data = top_id, label = top_id$ID, size = 2, label.size = NA, fill = NA, max.overlaps = Inf)
p3
```

### Apply t-test and Calculate Fold Change for Test Group 4 (NY_Post-vac)
```{r}
p_value = apply(df, 1, t_test_function, control = c(1:3), test_group = c(22:27))
test = apply(log2_df[,22:27], 1, mean, na.rm=TRUE)
LogFC <- control - test
new_data <- data.frame(ID = data[4:92,1], LogFC, p_value)
new_data <- new_data %>%
  mutate(Expression = dplyr::case_when(LogFC >= 0 & p_value <= 0.05 ~ "Up-regulated",
                                       LogFC <= 0 & p_value <= 0.05 ~ "Down-regulated",
                                       TRUE ~ "Unchanged"))
top_id <- new_data %>%
  filter(Expression == "Up-regulated" | Expression == "Down-regulated")
```

### Create Volcano Plot for Test Group 4 (NY_Post-vac)
```{r}
p4 <- ggplot(new_data, aes(x = LogFC, y = -log(p_value, 10))) +
  geom_point(aes(color = Expression)) +
  ggtitle("Gene Expression in NY_Post-vac") +
  theme(plot.title = element_text(size = 24, hjust = 0.5)) +
  xlab(expression("log"[2]*"FC")) +
  ylab(expression("-log"[10]*"FDR")) +
  scale_color_manual(values = c("green", "grey50", "firebrick3")) +
  geom_label_repel(data = top_id, label = top_id$ID, size = 2, label.size = NA, fill = NA, max.overlaps = Inf)
p4
```

### Apply t-test and Calculate Fold Change for Test Group 5 (Kappa_Post-vac)
```{r}
p_value = apply(df, 1, t_test_function, control = c(1:3), test_group = c(28:33))
test = apply(log2_df[,28:33], 1, mean, na.rm=TRUE)
LogFC <- control - test
new_data <- data.frame(ID = data[4:92,1], LogFC, p_value)
new_data <- new_data %>%
  mutate(Expression = dplyr::case_when(LogFC >= 0 & p_value <= 0.05 ~ "Up-regulated",
                                       LogFC <= 0 & p_value <= 0.05 ~ "Down-regulated",
                                       TRUE ~ "Unchanged"))
top_id <- new_data %>%
  filter(Expression == "Up-regulated" | Expression == "Down-regulated")
```

### Create Volcano Plot for Test Group 5 (Kappa_Post-vac)
```{r}
p5 <- ggplot(new_data, aes(x = LogFC, y = -log(p_value, 10))) +
  geom_point(aes(color = Expression)) +
  ggtitle("Gene Expression in Kappa_Post-vac") +
  theme(plot.title = element_text(size = 24, hjust = 0.5)) +
  xlab(expression("log"[2]*"FC")) +
  ylab(expression("-log"[10]*"FDR")) +
  scale_color_manual(values = c("green", "grey50", "firebrick3")) +
  geom_label_repel(data = top_id, label = top_id$ID, size = 2, label.size = NA, fill = NA, max.overlaps = Inf)
p5
```

### Apply t-test and Calculate Fold Change for Test Group 6 (Delta_Post-vac)
```{r}
p_value = apply(df, 1, t_test_function, control = c(1:3), test_group = c(34:39))
test = apply(log2_df[,34:39], 1, mean, na.rm=TRUE)
LogFC <- control - test
new_data <- data.frame(ID = data[4:92,1], LogFC, p_value)
new_data <- new_data %>%
  mutate(Expression = dplyr::case_when(LogFC >= 0 & p_value <= 0.05 ~ "Up-regulated",
                                       LogFC <= 0 & p_value <= 0.05 ~ "Down-regulated",
                                       TRUE ~ "Unchanged"))
top_id <- new_data %>%
  filter(Expression == "Up-regulated" | Expression == "Down-regulated")
```

### Create Volcano Plot for Test Group 6 (Delta_Post-vac)
```{r}
p6 <- ggplot(new_data, aes(x = LogFC, y = -log(p_value, 10))) +
  geom_point(aes(color = Expression)) +
  ggtitle("Gene Expression in Delta_Post-vac") +
  theme(plot.title = element_text(size = 24, hjust = 0.5)) +
  xlab(expression("log"[2]*"FC")) +
  ylab(expression("-log"[10]*"FDR")) +
  scale_color_manual(values = c("grey50", "firebrick3")) +
  geom_label_repel(data = top_id, label = top_id$ID, size = 2, label.size = NA, fill = NA, max.overlaps = Inf)
p6
```